---
title: "Houston Blackout Analysis (2021)"
description: |
  This analysis will be based on remotely-sensed night lights data, acquired from the Visible Infrared Imaging Radiometer Suite (VIIRS) onboard the Suomi satellite.The task is to answer the questions: How many residential buildings were without power on 2021-02-16? Is there a socioeconomic metric that predicts being affected by the power outage?
author:
  - name: Cullen Molitor
    url: {}
date: 2021-12-03
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
categories:
  - R
  - Remote Sensing
  - Spatial Analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Purpose

This notebook examines the effects of the February 2021 winter storms in Texas which caused blackouts. This examine the extent of the blackout over Houston using data from the [VIIRS](https://www.jpss.noaa.gov/viirs.html) instrument on the Suomi NPP satellite mission.

## Load libraries 

```{r loading packages}
library(sf)
library(tidyverse)
library(here)
library(stars)
library(rosm)
library(tmap)
```

## Function to load the DNB dataset from VNP46A1 granules

```{r read_dnb function}
read_dnb <- function(file_name) {
  # Reads the "DNB_At_Sensor_Radiance_500m" dataset from a VNP46A1 granule into a STARS object.
  # Then read the sinolsoidal tile x/y positions and adjust the STARS dimensions (extent+delta)

  # The name of the dataset holding the nightlight band in the granule
  dataset_name <- "//HDFEOS/GRIDS/VNP_Grid_DNB/Data_Fields/DNB_At_Sensor_Radiance_500m"

  # From the metadata, we pull out a string containing the horizontal and vertical tile index
  h_string <- gdal_metadata(file_name)[199]
  v_string <- gdal_metadata(file_name)[219]
  
  # We parse the h/v string to pull out the integer number of h and v
  tile_h <- as.integer(str_split(h_string, "=", simplify = TRUE)[[2]])
  tile_v <- as.integer(str_split(v_string, "=", simplify = TRUE)[[2]])

  # From the h/v tile grid position, we get the offset and the extent
  west <- (10 * tile_h) - 180
  north <- 90 - (10 * tile_v)
  east <- west + 10
  south <- north - 10

  # A tile is 10 degrees and has 2400x2400 grid cells
  delta <- 10 / 2400

  # Reading the dataset
  dnb <- read_stars(file_name, sub = dataset_name)

  # Setting the CRS and applying offsets and deltas
  st_crs(dnb) <- st_crs(4326)
  st_dimensions(dnb)$x$delta <- delta
  st_dimensions(dnb)$x$offset <- west
  st_dimensions(dnb)$y$delta <- -delta
  st_dimensions(dnb)$y$offset <- north
  
  return(dnb)
}
```

## Read in day night band (DNB) data
```{r read data}
Feb_07_v5 <- read_dnb(file_name = "data/VNP46A1.A2021038.h08v05.001.2021039064328.h5")
Feb_07_v6 <- read_dnb(file_name = "data/VNP46A1.A2021038.h08v06.001.2021039064329.h5")
Feb_16_v5 <- read_dnb(file_name = "data/VNP46A1.A2021047.h08v05.001.2021048091106.h5")
Feb_16_v6 <- read_dnb(file_name = "data/VNP46A1.A2021047.h08v06.001.2021048091105.h5")
```
## Combine adjacent tiles for each date

```{r combine data}
Feb_07_v5_v6 <- st_mosaic(Feb_07_v5, Feb_07_v6)
Feb_16_v5_v6 <- st_mosaic(Feb_16_v5, Feb_16_v6)
```

```{r remove unused objects and clean memory, include=FALSE}
rm(Feb_07_v5, Feb_07_v6, Feb_16_v5, Feb_16_v6)
gc()
```

## Create blackout Mask

```{r creating blackout mask}
## Take the difference of light data from before the storm and after the storm to make a mask of values with a difference greater than 200 nW cm-2 sr-1
diff <- (Feb_07_v5_v6 - Feb_16_v5_v6) > 200

## Convert values with a difference of less than 200 to NA
diff[diff == F] <- NA
```

## Vectorize blackout mask

```{r Vectorizing blackout mask}
blackout_mask <- st_as_sf(diff)

## Fix invalid geometries 
blackout_mask_fixed <- st_make_valid(blackout_mask)

rm(diff, blackout_mask)
gc()
```

## Crop the vectorized blackout mask to the region of interest

```{r croppping ROI}
## Set region of interest
houston <- st_polygon(
  list(
    rbind(
      c(-96.5, 29), 
      c(-96.5, 30.5), 
      c(-94.5, 30.5), 
      c(-94.5, 29), 
      c(-96.5, 29)
    )
  )
) %>% 
  st_sfc(crs = 4326)

## Crop night lights data
intersects <- st_intersects(blackout_mask_fixed, houston, sparse = FALSE)
blackout_cropped <- blackout_mask_fixed[intersects,]

## Transform cropped blackout mask back to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area)
blackout_cropped_NAD83 <- st_transform(blackout_cropped, 3083)

rm(blackout_cropped)
gc()
```

## Sanity check plot

```{r sanity check plot}
ggplot() +
  geom_sf(data = blackout_cropped_NAD83) +
  theme_classic()
```

## Roads data

```{r read in roads data with query}
query <- 
  "SELECT * 
   FROM gis_osm_roads_free_1 
   WHERE fclass='motorway'"

highways <- 
  st_read(
    "data/gis_osm_roads_free_1.gpkg", 
    query = query) %>% 
  st_transform(crs = 3083) %>% 
  st_buffer(dist = 200) %>% 
  st_union()
cat("\n\n\nAfter Transforming\n\n")
highways
```

## Basic highways plot

```{r}
ggplot() +
  geom_sf(data = highways) +
  theme_classic()
```

## Buildings data

```{r read in buildings data with query}
query <- 
  "SELECT * 
   FROM gis_osm_buildings_a_free_1
   WHERE (type IS NULL AND name IS NULL)
   OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')"

buildings <- 
  st_read(
    "data/gis_osm_buildings_a_free_1.gpkg", 
    query = query) %>% 
  st_transform(crs = 3083) 
cat("\n\n\nAfter Transforming\n\n")
buildings
```

## Census data

```{r read in census data }
# st_layers("ACS_2019_5YR_TRACT_48_TEXAS.gdb"))

acs_geoms <- 
  st_read(
    "data/ACS_2019_5YR_TRACT_48_TEXAS.gdb",
    layer = "ACS_2019_5YR_TRACT_48_TEXAS"
  )

acs_income <- 
  st_read(
    "data/ACS_2019_5YR_TRACT_48_TEXAS.gdb",
    layer = "X19_INCOME"
  ) %>% 
  select(GEOID, B19013e1) %>% 
  rename(GEOID_Data = GEOID,
         median_income = B19013e1)
```

```{r}
acs_geoms_med <- left_join(acs_geoms, acs_income)  %>% 
  st_transform(crs = 3083) 
```

## Merge datasets

```{r}
blackout_no_hwy <- st_difference(blackout_cropped_NAD83, highways)
rm(highways)
gc()
```

```{r}
houston_res_wo_power <- buildings[blackout_no_hwy, op = st_intersects]
number_houses_wo_power <- length(houston_res_wo_power$osm_id)
```

## The number of buildings that were left without power is `r number_houses_wo_power`.

```{r}
acs_building <- st_join(houston_res_wo_power, acs_geoms_med, join = st_intersects)
```

```{r}
acs_polygon <- st_join(blackout_no_hwy, acs_geoms_med, join = st_intersects)
```

```{r}
houston_bbox <- st_bbox(houston)
houston_map <- osm.raster(houston_bbox)
```

## Area and Median Incomes of Residences Affected by Houston Blackout in February, 2021 

```{r preview=TRUE}
tm_shape(houston_map) +
  tm_rgb(alpha = .75) +
  # tm_shape(acs_geoms_med,
  #          border.alpha = 0) +
  # tm_polygons(col = "median_income") +
  tm_shape(blackout_no_hwy) +
  tm_polygons(border.alpha = 1) +
  tm_shape(acs_polygon) + 
  tm_fill("median_income", 
          n = 5, 
          style = "pretty",
          title = "Median Income ($)") +
  tm_compass()+
  tm_scale_bar()
```

This map was created by Amber McEldowney and Cullen Molitor 2021-10-24.  <br> **Sources**:  Socioeconomic data: U.S. Census Bureau’s American Community Survey for Texas census tracts in 2019 <br> Light data:  NASA’s Level-1 and Atmosphere Archive & Distribution System Distributed Active Archive Center (LAADS DAAC) <br> Spatial & Buildings Data: OpenStreetMap.

```{r}
rm(blackout_cropped_NAD83, acs_geoms, acs_income, buildings)
gc()
```


## Histogram of Median Income for Houses Affected by Blackout

```{r}
Median_Income <- acs_polygon$median_income
hist(Median_Income,
main="Median Income in Regions Affected by Blackout",
xlab="Median Income ($)",
ylab="Number of Households",
col="grey",
freq=FALSE
)

```

## Histogram of Median Income in Houston

```{r}

houston_nad <- houston %>%
  st_as_sf() %>%
  st_transform(houston, crs = 3083)

# acs_geoms_med_bb <- st_join(acs_geoms_med, houston_nad, join = st_intersects)

intersects <- st_intersects(acs_geoms_med, houston_nad, sparse = FALSE)
acs_geoms_med_bb <- acs_geoms_med[intersects,]

Median_Income_Houston <- acs_geoms_med_bb$median_income
hist(Median_Income_Houston,
main="Median Income in Houston",
xlab="Median Income ($)",
ylab="Number of Households",
col="grey",
freq=FALSE
)
```

We thought it would be interesting to compare the median incomes of households affected by the blackout, to median incomes in Houston in general, but because the blackouts seem to have occured in a more metropolitan area, it is likely the incomes are skewed higher for that area, and it does not give a clear indication of whether median income had an effect on whether or not a household experienced a blackout.

