---
title: "Tidy Data"
description: |
  Some concepts and functions around `Tidy` data
author:
  - name: Cullen Molitor
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Today we introduced the `tidy` data concept where each column is a variable, each row is an observation, and each cell is a unique value. This format gives us a standardized way to quickly munge our data. The `tidyverse` suite of packages are designed to use messy raw data to create tidy data, through a standard set of arguments and syntax in the provided functions. `pivot_longer()` is a function to take wide format data and return long format tidy data. 

Here is a wide format dataset

```{r}
knitr::kable(dplyr::slice(mtcars, 1:10))
```

that we then make it into a long format dataset with `pivot_longer()`

```{r}
cars_long <- mtcars |> 
  tibble::rownames_to_column("car") |> 
  tidyr::pivot_longer(-1, names_to = 'var', values_to = 'value') |> 
  dplyr::slice(1:10)
knitr::kable(cars_long)
```

